---
title: "Building HTA 2.0 annotation package"
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 5
bibliography: references.bib

---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
options("citation_format" = "pandoc")

platformid="hta.2.0"
platformname="GeneChip® Human Transcriptome Array 2.0"
analystname="Andreas Sjodin"
analystemail="sjodin@hsph.harvard.edu"


library(knitr)
opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE, echo=FALSE,cache=FALSE, tidy.opts=list(keep.blank.line=FALSE, width.cutoff=120), dev="svg")
options(width=200)
```

---

Building annotation package for Affymetrix HTA 2.0 array: (`r platformname`)
Contact `r analystname` (`r analystemail`) for additional details.

The most recent update of this html document occurred: `r date()`

The sections below provide code to reproduce the building the transcriptcluster (gene) annotation package. 

---

# Setup

## Variables
Working directories, files and other variables necessary to the analysis.

```{r variables}
## Setup Data and Results directory variables
if(file.exists("/n/hsphS10/hsphfs1/chb/projects/mc_cordblood/")){
  baseDir="/n/hsphS10/hsphfs1/chb/projects/mc_cordblood/"
    } else if (file.exists("/Users/andreassjodin/hbc-projects/mc_cordblood")){
    baseDir="/Users/andreassjodin/hbc-projects/mc_cordblood"
    }

dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")


```

## Libraries

[Bioconductor](http://www.bioconductor.org) and [R](http://cran.r-project.org/) libraries used to process build annotation package.

```{r librariesvariables, echo=TRUE}
library(knitr) # for simple tables
library(AnnotationForge)

```

## Functions

```{r functions, echo=TRUE}
parseAffyTranscriptCsv <- function(probefile, orgpkg, fileout, headercol = "mrna_assignment"){ 
  dat <- read.csv(probefile, comment.char = "#", stringsAsFactors=FALSE,  na.string = "---")
  mrna <- sapply(strsplit(dat[,headercol], " // | /// "), function(x)  grep("^[NX][MR]|^[A-G][A-Z]+[0-9]+|^[A-Z][0-9]+|^ENST", x, value = TRUE)[1])
  
  ens <- grep("^ENS", mrna, value = TRUE)
  require(orgpkg, character.only = TRUE) || stop(paste("You need to install the", orgpkg, "package first!"))
  ens <- select(get(orgpkg), ens, c("REFSEQ","ACCNUM"), "ENSEMBLTRANS")
  ens <- ens[!duplicated(ens[,1]),]
  
  ## use accnum if refseq is NA
  ens[is.na(ens[,2]),2] <- ens[is.na(ens[,2]),3]
  
  ## put mapped data back in mrna vector
  mrna[match(ens[,1], mrna)] <- ens[,2]
  mrna[grep("^ENS|^GENSCAN", mrna)] <- NA
  
  ## write out
  write.table(cbind(dat[,1], mrna), fileout, sep = "\t", quote = FALSE, row.names = FALSE, col.names  = FALSE, na = "")

}
```

---

# Building package


Set up variables
```{r annotationvariables, echo=TRUE}
affytranscriptfile <- "annotation_package/HTA-2_0.na34.hg19.transcript.csv"
orgdb <- "org.Hs.eg.db"
mapperfile <- "HTA20_mapper.txt"
dbschema <- "HUMANCHIP_DB"
fileprefix <- "hta20transcriptcluster"
annoversion <- "1.0.0"
chipname <- "Human Transcriptome Array 2.0"
```

Create file needed for AnnotationForge (transcript id and Entrez gene)

```{r parsedata, echo=TRUE}
parseAffyTranscriptCsv(affytranscriptfile, orgdb, mapperfile)
```


Create annotation package

```{r loadid, results='hide', echo=TRUE}

makeDBPackage(
schema=dbschema,
affy=FALSE,
prefix=fileprefix,
fileName=mapperfile,
baseMapType="gbNRef",
outputDir = getwd(),
version= annoversion,
manufacturer = "Affymetrix",
chipName = chipname,
manufacturerUrl = "http://www.affymetrix.com")


```


Install created annotation package
```{r installpackage, echo=TRUE, eval=FALSE}
install.packages(paste(fileprefix, ".db", sep=""), repos=NULL, type="source")

```



---

# R Session Info

(useful if replicating these results)

```{r sessioninfo}
sessionInfo()
```

